---
title: "R Markdown and Leaflet Assignment"
author: "Kannan Govindarajan"
date: "December 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

## 23 Great Schools with Master's Programs in Data Science

This assignment provides interactive map of 23 great schools which offer Master's degree program in Data Science with hyperlink pointing to the details of the program.

Universities Data: 
http://www.mastersindatascience.org/schools/23-great-schools-with-masters-programs-in-data-science/

Universities Address Data:
http://nces.ed.gov/ipeds/datacenter/data/HD2015.zip

Latitude Longitude Data:
Google Maps API - R package ggmap.geocode(address)

## Load Data

Universities data offering data science masters are directly read from the web page  using RCurl and necessary data is extracted using XML package.
```{r}
library(bitops)
library(RCurl)
library(XML)

url <- "http://www.mastersindatascience.org/schools/23-great-schools-with-masters-programs-in-data-science/"
html <- getURL(url)
doc <- htmlParse(html, asText=TRUE)
universityNames <- xpathSApply(doc, "//ul[@class='top-23-list']/li/h3", xmlValue)
courseURL <- xpathSApply(doc, "//ul[@class='top-23-list']/li/h3/a", xmlGetAttr, 'href')


```

Universities address data are loaded from the source http://nces.ed.gov.
```{r}
temp <- tempfile()
download.file("http://nces.ed.gov/ipeds/datacenter/data/HD2015.zip",temp, mode="wb")
unzip(temp, "hd2015.csv")
dd <- read.table("hd2015.csv", sep=",",header=T)
colnames(dd)[2]<-"universityNames"

dd$Address <- paste(dd$ADDR,",",dd$CITY, ",",dd$STABBR)

```

Merged data from above two sources, by matching the university names.
```{r}
universityNames[universityNames == 'Arizona State University'] <- "Arizona State University-Tempe"
universityNames[universityNames == 'Columbia University'] <- "Columbia University in the City of New York"
universityNames[universityNames == 'Georgia Tech'] <- "Georgia Institute of Technology-Main Campus"
universityNames[universityNames == 'North Carolina State University'] <- "North Carolina State University at Raleigh"
universityNames[universityNames == 'Rutgers University'] <-"Rutgers University-Camden"
universityNames[universityNames == 'Texas A&M University'] <-"Rutgers University-Camden"
universityNames[universityNames == 'University of Virginia'] <- "University of Virginia-Main Campus"
universityNames[universityNames == 'University of Wisconsin'] <- "University of Wisconsin-Madison"
universityNames[universityNames == 'Texas A&M University'] <- "Texas Southern University"
universityNames <- gsub(", ","-",universityNames)

df <- cbind(universityNames, courseURL)

lData <- merge(df,dd[,c("universityNames","Address")])

```

Universities address data is supplied to Google maps API to fetch the lattitude and longitude data for display using Leaflet.
```{r message=FALSE}
library(ggplot2)
library(ggmap)

latlong <- geocode(lData$Address, output = "latlona",messaging = FALSE)
lData$lat <- latlong$lat
lData$long <- latlong$lon

```

## Display Interactive Map

Display the data as interactive map using Leaflet.
```{r}
library(leaflet)

lData <- na.omit(lData)

leaflet(lData) %>% addTiles() %>% addMarkers(~long, ~lat, popup = paste("<a href=",lData$courseURL,">", lData$universityNames,"</a>"), clusterOptions = markerClusterOptions())

```



